Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-31 _tex/paper2.tex
Built site for gh-pages

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues that can undermine the reliability and validity of CJ's
trait estimates. The first issue arises from the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
The second stems from the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing. The study then addresses these
issues by extending Thurstone's model through a systematic and
integrated approach based on causal and Bayesian inference methods.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's model to
address these challenges. The model integrates the theoretical core
principles alongside key assessment design features relevant to CJ
experiments, such as the selection of judges, stimuli, and comparisons.
Section~\ref{sec-statistical} translates these theoretical and practical
elements into a probabilistic statistical model for the analysis of
pairwise comparison data. Section~\ref{sec-discussion} discusses the
implications of the findings and explores avenues for future research.
Finally, Section~\ref{sec-conclusion} summarizes the study's key
insights.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons. A thought
experiment can illustrate this concept. In this experiment, the
researcher observes the discriminal processes for the texts depicted in
Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
assumes that the discriminal dispersion for Text A remains constant and
that the texts are uncorrelated \((\rho=0)\). In this scenario,
Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
associated with the perception of Text B in comparison to Text A,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B > A)\), ultimately influencing the comparison
outcome. Additionally, the figure reveals that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more narrow compared to situations where their
dispersions differ. Consequently, the discriminal difference is more
likely to favor Text B over Text A (shaded gray area)

@@

In aggregate, the overreliance on Thurstone's Case V assumptions in the
statistical analysis of CJ data can compromise the reliability of CJ's
trait estimates. This overreliance, coupled with the apparent disconnect
between CJ's approach to trait measurement and hypothesis testing, may
also undermine the validity of these estimates
\citep[pp.~2]{Perron_et_al_2015}. Consequently, addressing these issues
by adopting a more systematic and integrated approach could offer
several statistical and measurement benefits.

@@

For instance, researchers might aim to answer the question: ``To what
extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, they could
design a CJ experiment by randomly assigning students to two groups,
each receiving a different teaching method. Judges would then compare
pairs of students' written texts, producing a dichotomous outcome that
reflects the relative quality of each text. In this scenario,
researchers can reformulate the research question as the estimand:
``\emph{On average}, is there a difference in the ability to produce
high-quality written texts between the two groups of students?''.
Researchers would then use estimates from the BTL model, or its
transformations, to approximate this estimand.

However, Section~\ref{sec-theory-issues} presents strong evidence that
the BTL model has several statistical and measurement limitations. These
limitations hinder the model's ability to identify distinct estimands
relevant to CJ inquiries, including the one presented in the previous
paragraph. Identification is crucial because it is a necessary condition
for ensuring consistent estimators. Without identification, researchers
cannot achieve consistency even with ``infinite'' and error-free data.
In other words, the estimates derived from the estimator fail to
converge to the ``true'' value of the estimand as the data size
approaches infinity \citep{Everitt_et_al_2010}. As a result, deriving
meaningful insights from finite data also becomes unattainable
\citep[pp.~5]{Schuessler_et_al_2023}.

@@

As a result, this section addresses the identified limitations by
extending Thurstone's model through a systematic and integrated approach
based on the structural approach to causal inference. Specifically, it
leverages the capabilities of this approach to formalize the combination
of the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments. These features include the selection of
judges, stimuli, and comparisons. Combining these components clarifies
the interactions among all actors and processes involved in CJ
experiments. Accordingly, Section~\ref{sec-theory-theoretical_P}
incorporates the theoretical principles into a
\emph{conceptual-population model}. Meanwhile,
Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
features into a \emph{sample-comparison model}.

@@

Before integrating the theoretical principles from
Section~\ref{sec-thurstone_theory}, it is important to define SCMs and
DAGs. SCMs are formal mathematical models defined by a set of
\emph{endogenous} variables, a set of \emph{exogenous} variables, and a
set of functions \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other variables. These functions use the
symbol `\(:=\)' to denote the asymmetrical causal dependence between
variables and the symbol `\(\:\bot\:\)' to represent
\emph{d-separation}, a concept akin to (conditional) independence.

For instance, following the scenario outline in
Section~\ref{sec-theoretical}, Figure~\ref{fig-cj04_scm} displays the
SCM that relates the outcome of the CJ experiment \((O_{abhijk})\), to
several endogenous variables \(V\). These variables include the
discriminal difference \((D_{abhijk})\), perceived discriminal processes
\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\), formally
expressed as
\(V = \{D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
The structural equations \(\{f_{O},f_{D},f_{PT}\}\) define the
dependencies among these variables. Notably, in this scenario, the
judges produce a dichotomous outcome \((O_{abhijk}) = \{0,1\}\).
Moreover, the indices \(\{a,b\}\) represent the compared texts,
\(\{h,i\}\) label the students who wrote these texts, \(j\) denotes the
judge, and \(k\) represents the comparison number. Regarding the \(k\)
index, the index accounts for cases in which judges perform multiple
comparisons of the same texts written by the same students, as the one
discussed in Section~\ref{sec-theoretical}.

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where the nodes represent random
variables. The term \emph{directed} means that the edges extend from one
node to another, with arrows indicating the direction of causal
influence. The term \emph{acyclic} implies that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} includes them, offering offering the
advantage of highlighting potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}.

@@

Because it quickly becomes intractable, and without loosing generality,
the (latent) ``perceived'' and ``true'' discriminal processes for the
stimuli can be depicted in a vector for each judge, as in

@@

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) is thought to be a confounder in the relationship between
the teaching method \(T\) and the outcome \(Y\). In this scenario, the
figures display multiple descendant relationships such as
\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\). They
also highlight unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).

>>>>

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues that can undermine the reliability and validity of CJ's
trait estimates. The first issue arises from the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
The second stems from the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing. The study then addresses these
issues by extending Thurstone's general form through a systematic and
integrated approach based on causal and Bayesian inference methods.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The new model integrates the
theoretical core principles alongside key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. Section~\ref{sec-statistical} translates these
theoretical and practical elements into a probabilistic statistical
model for the analysis of pairwise comparison data.
Section~\ref{sec-discussion} discusses the implications of the findings
and explores avenues for future research. Finally,
Section~\ref{sec-conclusion} summarizes the study's key insights.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons. A thought
experiment can illustrate this concept. In this experiment, the
researcher observes the discriminal processes for the texts depicted in
Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
assumes that the discriminal dispersion for Text A remains constant and
that the texts are uncorrelated \((\rho=0)\).
Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
associated with the perception of Text B in comparison to Text A,
\((\sigma_{B}-\sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B > A)\), ultimately influencing the comparison
outcome. Additionally, the figure reveals that when the discriminal
dispersions of the texts are equal \((\sigma_{B}-\sigma_{A}=0)\), the
discriminal difference is more narrow compared to situations where their
dispersions differ. Consequently, the discriminal difference is more
likely to favor Text B over Text A (shaded gray area)

@@

In aggregate, the excessive reliance on Thurstone's Case V assumptions
in the statistical analysis of CJ data can compromise the reliability of
CJ's trait estimates. This overreliance, coupled with the apparent
disconnect between CJ's approach to trait measurement and hypothesis
testing, may also undermine the validity of these estimates
\citep[pp.~2]{Perron_et_al_2015}. Thus, addressing these issues through
a more systematic and integrated approach could provide significant
statistical and measurement benefits. For this reason, the study employs
the structural approach to causal inference, as it fulfills these
requirements.

@@

For instance, a motivating example that illustrates these concepts
throughout the document could involve researchers aiming to answer the
question: ``To what extent do different teaching methods influence
students' ability to produce high-quality written texts?'' To
investigate this, they could design a CJ experiment by randomly
assigning students to two groups, each receiving a different teaching
method. Judges would then compare pairs of students' written texts,
producing a dichotomous outcome that reflects the relative quality of
each text. Based on this setup, researchers can reformulate the research
question as the estimand: ``\emph{On average}, is there a difference in
the ability to produce high-quality written texts between the two groups
of students?''. Following current CJ practices, researchers would then
rely on estimates from the BTL model, or its transformations, to
approximate this estimand.

However, Section~\ref{sec-theory-issues} presents compelling evidence
that the BTL model has several statistical and measurement limitations.
These limitations hinder the model's ability to identify distinct
estimands relevant to CJ inquiries, including the one from the
motivating example. Identification is crucial because it is a necessary
condition for ensuring consistent estimators. Without identification,
researchers cannot achieve \emph{consistency} even with ``infinite'' and
error-free data. In other words, the estimates derived from the
estimator would fail to converge to the ``true'' value of the estimand
as the data size approaches infinity \citep{Everitt_et_al_2010}. As a
result, deriving meaningful insights from finite data also becomes
unattainable \citep[pp.~5]{Schuessler_et_al_2023}.

@@

As a result, this section addresses the identified limitations by
extending Thurstone's general form (see Table~\ref{tbl-thurstone_cases})
through a systematic and integrated approach based on the structural
approach to causal inference. Specifically, it leverages the
capabilities of this approach to formalize the combination of the core
theoretical principles outlined in Section~\ref{sec-thurstone_theory}
with key assessment design features relevant to CJ experiments, such as
the selection of judges, stimuli, and comparisons. The extended model
clarifies the interactions among all actors and processes involved in CJ
experiments. Furthermore, it allows researchers to shift the current
comparative data analysis paradigm from accepting the model's
assumptions as given to testing these assumptions for specific
comparative data. For details on these assumptions, refer to
Section~\ref{sec-theory-issues}. Accordingly,
Section~\ref{sec-theory-theoretical_P} incorporates the theoretical
principles into what the study calls the \emph{conceptual-population
model}. Meanwhile, Section~\ref{sec-theory-theoretical_SC} integrates
the assessment design features into what the study terms the
\emph{sample-comparison model}.

@@

Before integrating the theoretical principles from
Section~\ref{sec-thurstone_theory}, the study defines SCMs and DAGs.
SCMs are formal mathematical models characterized by a set of
\emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
\(E\), and a set of functions \(F\)
\citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous variables are those
whose causal mechanisms a researcher chooses to model \citep{Neal_2020}.
In contrast, exogenous variables represent \emph{errors} or
\emph{disturbances} arising from omitted factors that the investigator
chooses not to model explicitly \citep[pp.~27,68]{Pearl_2009}. Lastly,
the functions, referred to as \emph{structural equations}, express the
endogenous variables as non-parametric functions of other variables.
These functions use the symbol `\(:=\)' to denote the asymmetrical
causal dependence between variables and the symbol `\(\:\bot\:\)' to
represent \emph{d-separation}, a concept akin to (conditional)
independence.

Figure~\ref{fig-cj04} illustrates the preliminary integration of the
theoretical principles underlying the motivating example. Specifically,
SCM~\ref{fig-cj04_scm} displays the relationship between the outcome of
the CJ experiment, \(O_{abhijk}\), to several other variables. Here, the
indices \(\{a,b\}\) represent the compared texts, \(\{h,i\}\) label the
students who wrote these texts, \(j\) denotes the judge, and \(k\)
accounts for cases where the judges perform multiple comparisons of the
same texts written by the same students, i.e., the comparison number for
a \emph{repeated measures design} \citep[pp.~366-376]{Lawson_2015}. The
other variables related to the outcome include the discriminal
difference \((D_{abhijk})\), the perceived discriminal processes
\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\). Together
with the outcome, these variables form the preliminary set of endogenous
variables
\(V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).
Finally, the preliminary set of structural equations
\(F = \{f_{O},f_{D},f_{PT}\}\) defines the dependencies among these
variables.

@@

Notably, every SCM has an associated DAG
\citep{Pearl_et_al_2016, Cinelli_et_al_2020}. A DAG is a graph
consisting of nodes connected by edges, where the nodes represent random
variables. The term \emph{directed} means that the edges extend from one
node to another, with arrows indicating the direction of causal
influence. The term \emph{acyclic} implies that the causal influences do
not form loops, ensuring the influences do not cycle back on themselves
\citep{McElreath_2020}. DAGs represent observed variables as solid black
circles, while they use open circles for unobserved (latent) variables
\citep{Morgan_et_al_2014}. While the \emph{standard representation} of
DAGs typically omits exogenous variables for simplicity, the
\emph{magnified representation} includes them, offering the advantage of
highlighting potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}.

DAG~\ref{fig-cj04_dag} displays the preliminary DAG associated with
SCM~\ref{fig-cj04_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory. It also distinguishes between observable endogenous variables,
such as the outcome \(O_{abhijk}\), and variables considered latent
according to the theory and practice of CJ, such as the discriminal
difference \((D_{abhijk})\), the perceived discriminal processes
\((PT_{ahjk}, PT_{bijk})\), the ``true'' discriminal processes
\((T_{ahjk}, T_{bijk})\), and the judges' biases \((B_{kj})\).

@@

Researchers can then use these building blocks to represent the scenario
outlined in Section~\ref{sec-appendix-B}. SCM \ref{fig-scm_example1} and
DAG \ref{fig-mdag_example1} depict the plausible causal structure for
this example. In this context, the variable \(X\) (socio-economic status
of the school) is thought to be a confounder in the relationship between
the teaching method \(T\) and the outcome \(Y\). The figures display
multiple descendant relationships such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). They also highlight
unconnected node pairs, evident from the relationships
\(e_{T} \:\bot\:e_{X}\), \(e_{T} \:\bot\:e_{Y}\), and
\(e_{X} \:\bot\:e_{Y}\). Additional, the figures show one fork,
\(X \rightarrow \{T, Y\}\), and two colliders:
\(\{X, e_{T}\} \rightarrow T\) and \(\{X, T, e_{Y}\} \rightarrow Y\).
